---
title: "`BulkSignalR extension` : </br> Integration of post-translational modifications for inference of ligand-receptor interactions"
author:
  - name: Pierre Giroux
affiliation:
  - Institut de Recherche en Cancérologie de Montpellier, Inserm, Montpellier, France 
email: pierre.giroux@inserm.fr
- name: Jacques Colinge
affiliation:
  - Institut de Recherche en Cancérologie de Montpellier, Inserm, Montpellier, France 
email: jacques.colinge@inserm.fr
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output: 
  rmarkdown::html_vignette:
  self_contained: true
toc: true
toc_depth: 4
highlight: pygments
fig_height: 3
fig_width: 3
fig_caption: no
code_folding: show
vignette: >
  %\VignetteIndexEntry{BulkSignalR-PTM}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE, 
  comment = "##"
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

BulkSignalR extension package version: `r packageVersion("PTMSignalR")`

## PTM mode 

This extension of `BulkSignalR` allows the user to request different pairwise
comparisons from a large cohort of samples and to integrate post-translational 
modifications associated to expression data.

In this mode, ligand-receptor interactions are inferred based on gene or protein
regulation-associated P-values when comparing two clusters of samples. 
Post-translational modifications are identified from target genes of pathways 
predicted using expression data. P-values are obtained comparing two clusters of
samples.

The user must first perform a differential expression analysis 
(using the tool of his choice as DESeq2, EdgeR...) between a 
pair of sample clusters. Differential analysis of post-translational 
modifications data must be performed between same pair of clusters.


In the next chunk of code, we describe an application to 
Clear Cell Renal Cell Carcinoma (ccrcc) samples where we compare two clusters 
of patients.

```{r diffmode,eval=FALSE}
library(PTMSignalR)

data(ccrcc,package='PTMSignalR')

protResp <- clinical[clinical$Response=="Responder",2]
protNoResp <- clinical[clinical$Response=="Non-Responder",2]
phosResp <- clinicalPhos[clinicalPhos$Response=="Responder",2]
phosNoResp <- clinicalPhos[clinicalPhos$Response=="Non-Responder",2]

# Normalization by protein expression 
dfPhosNorm <- as.data.frame(dfPhos[dfPhos$Symbol %in% rownames(dfProt),])
tmp <- rownames(dfPhosNorm)
dfPhosNorm_tmp <- dfPhosNorm[,-c(1,2)]
for(i in 1:nrow(dfPhosNorm_tmp)){
  j <- dfPhosNorm$Symbol[i]
  dfPhosNorm_tmp[i,] <- dfPhosNorm_tmp[i,]/dfProt[j,]
}
dfPhosNorm_tmp[sapply(dfPhosNorm_tmp, is.infinite)] <- NA
dfPhosNorm <- cbind(dfPhosNorm[,c(1,2)], dfPhosNorm_tmp)
rownames(dfPhosNorm) <- tmp


# Differential expression analysis for proteins
diff.resp <- apply(dfProt,1, function(x) wilcox.test(x[unlist(phosResp)],
                                                x[unlist(phosNoResp)])$p.value)
FC <- apply(dfProt,1,function(x) median(x[unlist(phosResp)], 
                                na.rm=T)-median(x[unlist(phosNoResp)], na.rm=T))
resp.table <- data.frame(pval=diff.resp,logFC=FC)
rownames(resp.table) <- rownames(dfProt)

# Differential expression analysis for phosphorylation
dfPhos_complete <- dfPhosNorm
dfPhosNorm <- dfPhos_complete[,-c(1,2)]
diff.respPh <- apply(dfPhosNorm,1, function(x) wilcox.test(x[unlist(phosResp)],x[unlist(phosNoResp)])$p.value)
FCPh <- apply(dfPhosNorm,1,function(x) median(x[unlist(phosResp)], 
                                na.rm=T)-median(x[unlist(phosNoResp)], na.rm=T))
resp.tablePh <- data.frame(pval=diff.respPh,logFC=FCPh)
rownames(resp.tablePh) <- rownames(dfPhosNorm)

bsrdmResp <- prepareDataset(counts = dfProt, phospho = dfPhos_complete, 
                    single = T, symbolPh.col = 1, pos.col = 2, normalize = T, method="TC",min.count=0,min.LR.found=10)
bsrdm.compResp <- as.BSRDataModelCompPhospho(bsrdmResp)

# indices resp vs non resp
indPhosResp <- which(colnames(dfPhos) %in% unlist(phosResp))
indPhosNoResp <- which(colnames(dfPhos) %in% unlist(phosNoResp))

# define the comparison
bsrccResp <- defineClusterComp(bsrdm.compResp, indPhosResp, indPhosNoResp, 
                               resp.table, resp.tablePh)
bsrccResp

bsrdm.compResp <- addClusterCompPhos(bsrdm.compResp,bsrccResp,"resp.versus.all")
bsrdm.compResp

# Infer ligand-receptor interactions from the comparison
bsrinf.comp.resp <- initialInference(bsrdm.compResp,"resp.versus.all",
                  min.logFC=minFC,max.pval = 1,neg.receptors = T,rank.p = 0.75,
                  min.pw.size = 2,min.positive = 2, reference="REACTOME")
bsrinf.comp.resp
resp.inter <- LRinter(bsrinf.comp.resp)



```  

##  Technical notes

Three previously described S4 objects (BSR-DataModelComp, BSR-InferenceComp, BSR-ClusterComp) 
have been extented :  
  
  \


* **BSR-DataModelCompPTM**, denoted `bsrdm.comp.ptm` is an extension from 
**BSR-DataModelComp**, previously denoted `bsrdm.comp`  
* **BSR-InferenceCompPTM**, denoted `bsrinf.comp.ptm` is an extension from 
**BSR-InferenceComp**, previously denoted `bsrinf.comp` 
* **BSR-ClusterCompPTM**, denoted `bsrcc.ptm` is an extension from 
**BSR-ClusterComp**, previously denoted `bsrcc` 



This new design let us handle PTM data in addition of expression data.

\

Thank you for reading this guide and for using `BulkSignalR`.  

## Session Information

```{r session-info}
sessionInfo()
```
